Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-10 treestats_paper.tex
above/below -> ancestral/descending: closes #8

Many population genetic summary statistics can be defined using the allele frequency spectrum
of a collection of homologous sequences.
Concretely, write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
-- i.e., the number of alleles whose frequency is $p$ in our sample --
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci, as
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
(Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
occur below that mutation in the tree (and accounting for those that subsequently mutate).
Here, we only consider statistics whose values are invariant to monomorphic sites. Thus, for a sample of $n$
genomes, $0 < p \leq 1-1/n$.

What we do here is a generalization of this idea.
Roughly speaking, to do this we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics
(e.g., averaged in windows).
For instance, we could calculate how many of a certain set $A$ of samples
are below a particular edge in a tree
by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
(b) finding the total weight of all nodes in the subtree below that edge.
It turns out to be useful to allow more general weights (including negative ones).
Figure \ref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarising them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

@@

We allow vector-valued weights,
i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
so that when summarising we have access to more than one aspect of each subtree.
The requirement that $f(0) = f(\tiw) = 0$ is for consistency,
so that statistics do not depend on portions of the tree either not ancestral to any of the samples or ancestral to all
of them.
% Below, in section XXX, we describe how to efficiently
% maintain a list of the weights of the subtrees below each node
% as we iterate through the tree sequence.

\begin{figure}
    \centering
    \includegraphics{figures/divergence_diagram}
    \caption{
        An example of computing sequence divergence between red nodes and blue nodes
        from seven mutations at distinct sites in a sequence of length 10 using the tree.
        The summary function, $f(a,b) = (a/3)(1-b/2)$ gives the probability that
        an allele found in $a$ of the red nodes and $b$ of the blue nodes is present in
        a randomly chosen red node but not in a randomly chosen blue node.
        \textbf{(a)} Weights are assigned to each subtree by adding together the weights of any children,
        plus the weight of the node itself, if it is a sample.
        \textbf{(b)} In this example, subtree weights $(a,b)$ record how many red nodes ($a$)
        and how many blue nodes ($b$) lie in that subtree.
        \textbf{(c)} The summary function shown is applied to both the weight below
        and the weight not below each mutation,
        so for instance the mutation above the node with weights $(2,1)$
        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
        The final value for divergence is the sum of these values divided by 10,
        the total number of sites.
        \krt{This is a nice fig, but the $2 \times$ and the $/10 sites$ parts may be easy to miss?}
        \label{fig:divergence_diagram}
    }
\end{figure}

@@

Before defining the different classes of statistic,
it will help to establish a little more notation.
A tree sequence efficiently describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
% These trees are highly correlated,
% a fact which allows them to be stored and processed very efficiently,
% as described in \citet{kelleher2018efficient}.
From the tree sequence can be extracted a sequence of trees,
$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome betwen $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance the branch describing a relationship where $u$ is the child of $v$
is associated with $u$.
The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
and is denoted $t_T(u)$ (dropping $T$ if it is clear from context).
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships
can cause the precise set of samples that lie below that edge to differ across
trees due to recombination.

@@

Perhaps the most natural way to summarise weights on the tree
is simply to examine the values at each node.
This would allow us, for instance, to
count the number of samples that inherit from each node in the tree.
Averaged across trees,
this tells us what proportion of the sample's genomes were inherited from each node.
Motivated by this, we define the
\textbf{node statistic} for node $u$
associated with summary function $f()$ and sample weights $\iw$
to be the sum of $f()$ applied to the weight of the subtree below node $u$
and to the remaining weight,
averaged across the genome:
\begin{align}
    \node(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k \left( f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right).
\end{align}
The weight $\nw_{T_k}(u)$ is the subtree weight below $u$ in the tree $T_k$,
and $\tiw - \nw_{T_k}(u)$ is the sum of all weights \emph{not} in the subtree below $u$ in $T_k$.
% i.e., the length of the segment of genome that the tree extends for.
% If this is computed over only a ``window'' of the genome
% then $L_k$ is replaced by the length of the portion of that window
% that $T$ extends for.

@@

We will use weights that count number of samples below each node
sufficiently often that we give them a special name --
for a subset set $S$ of samples,
the \emph{indicator weights of $S$} are the sample weights $\bone_S$ with
$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.

\begin{example}[Ancestry proportions] \label{ex:ancestry_props}
    If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
    and then $\iw_u / n$ is the proportion of the samples in $S$ that lie below $u$.
    Therefore, if $f(x) = x / n$,
    then $\nodep(f, \iw)_u$ is the proportion of the genomes of $S$
    that are inherited from (ancestor) $u$.
\end{example}

@@

The summaries we defined above for each node in each tree
extend nicely to genetic variants
-- just as a node weight contains information about which samples are below the node and which are not,
so we can summarise patterns of genetic variation by summing up weights of all samples
that carry a given allele.
So, we define \emph{allele weights} to be
the total weight of all sample nodes who have inherited that allele.

@@

If there has been only one mutation at the site,
then $\aw_j(a)$ is equal to the weight of the subtree below the mutation that produced $a$.
More generally, $\aw$ is not necessarily equal to a subtree weight
but is easily computable from them.

@@

\begin{example}[Nucleotide diversity] \label{ex:site_diversity}
    The nucleotide diversity of a sample is the average number of differences between all haplotypes, which is
    equivalent to the sum of heterozygosity at all positions.
    To calculate this statistic,
    % let $\iw_u = 1$ for $u \in S$ and $\iw_v = 0$ otherwise,
    let $\iw = \bone_S$,
    so that $\nw(u)$ gives the number of nodes in $S$ below $u$,
    and define
    \begin{align*}
        f(x) = \frac{x (n - x)}{n (n-1)} .
    \end{align*}
    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide diversity in the region between $a$ and $b$.
\end{example}

\begin{example}[Nucleotide divergence] \label{ex:site_divergence}
    Now suppose we want to compute the mean density of nucleotide differences
    \emph{between} two nonoverlapping groups of samples, $S_1$ and $S_2$,
    with $n_1$ and $n_2$ samples, respectively.
    As before,
    let $\iw_j = \bone_{S_j}$,
    so that $\nw_{j}(u)$ gives the number of nodes in $S_j$ below $u$.
    Then we define
    \begin{align*}
        f(x_1, x_2) = \frac{x_1 (n_2 - x_2)}{n_1 n_2} .
    \end{align*}
    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide divergence between the two groups
    in the region between $a$ and $b$.
\end{example}

@@

This makes it natural to define
a statistic of tree shape by summing these expected contributions across its branches.
We define the \textbf{branch statistic} of a single tree $T$ with span $L$
obtained from summary function $f()$
and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
to be the sum of the area of each branch in each tree
multiplied by the summary function applied to its subtree weight:
\begin{align}
    \branch(f, \iw)_T
    &=
    \frac{1}{L} \sum_{u \in T} t_{u} \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
\end{align}
where $t_{u}$ is the length of the branch above node $u$ in tree $T$
and $\nw_{T}(u)$ is the total weight of the subtree of $T$ below $u$ (as defined above).
The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ below $u$,
because the sum of all weights in the tree is always $\tiw$.
The value $f(\nw(u))$ is the summary value that would be added to a site statistic
if a single mutation occurred on the branch above $u$,
and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
so $\branch(f, \iw)_T$ gives the expected contribution of this branch to $\site(f, \iw)$,
per unit of sequence length that the tree $T$ covers.

An example is shown in Figure~\ref{fig:branch_site_diagram},
showing how the $f_4$ site statistic assigns a weight to each mutation
depending on the frequencies in each of the four sample sets below it,
and how the corresponding branch statistic assigns the same weight to those branches.

@@

\begin{example}[Mean TMRCA] \label{ex:branch_diversity}
    If we take $\iw$ and $f$ exactly as in the ``Nucleotide diversity'' example above,
    then $f(u)$ gives the probability that the branch above $u$
    lies on the path from two randomly chosen samples from $S$
    on the path up to their most recent common ancestor (MRCA).
    Therefore, $\branch(f, \iw)$ for these choices
    gives the mean total distance in the tree between two samples from $S$,
    averaged across the sequence.
    This is twice the mean TMRCA if the samples are all from the same time.
\end{example}

@@

Thus far we have examined how statistics vary along the genome,
but a tree sequence allows us access to another dimension: time.
Next, we examine how the \emph{time} distribution of admixture signal
in the tree sequence inferred by tsinfer/GEVA.
The branch $f_4(A, B; C, D)$ statistic for four populations $A$, $B$, $C$, and $D$
sums $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and height,
where $p_A$ is the proportion of samples in $A$ that lie below that node, etcetera.
A branch with a large value is expected to contribute a large amount to the site $f_4$,
which we calculate from data and is used to detect historical connections between populations.
To examine the time signature, we assigned the contribution of each branch
to its terminal node,
and then summed these values across all nodes in 50 logarithmically-spaced bins by time.
We then normalised these to sum to 1 by dividing by the total branch $f_4$ value.

@@

\paragraph{Diploid statistics}
The statistics we develop here are in essence \emph{haploid} statistics.
These tools can be used to compute a statistic about polyploid individuals,
but they do not scale to large numbers of individuals.
For instance, to compute the heterozygosity of a set of samples
-- this is $(1/N) \sum_{i=1}^N 2 p_i (1-p_i)$, where $p_i$ is the derived allele frequency
in the $i^\text{th}$ individual --
one would need to have an entry in the weight vector for \emph{each individual}.
The genotype of a diploid at a site with a single mutation that occurred on the branch above node $n$
is determined by where that node falls relative to the diploid's two haplotypes in the tree:
if $n$ is on the path from either haplotype to their MRCA, the individual is heterozygous;
if $n$ is on the path from the MRCA to the root, the individual is homozygous derived;
and the individual is homozygous ancestral otherwise.
This suggests the following generalization of our current scheme for diploids:
for a list $(I_1, \ldots, I_N)$ of pairs of nodes,
and two vectors $w^{(1)}$ and $w^{(2)}$ of weights of length $N$,
add weight $w^{(1)}_i$ to each node on the path between the two nodes of $I_i$,
and weight $w^{(2)}_i$ to each node on the path from the MRCA of $I_i$ to the root.
With $k$ such weighting schemes, we would then need the summary function $F$
to take $k \times 2$ values as input.
Efficiently updating these ``diploid'' weights on the nodes should be possible,
but is not a straightforward extension of our current method.

@@

Let $G$ denote the genotype matrix, so that $G_{ja} \in \{0,1\}$
is the genotype of the $a^\text{th}$ individual
at the $j^\text{th}$ site, with `1` denoting the derived state
(we assume here biallelic markers).
Also let $P_j = (1/n) \sum_a G{ja}$ denote the vector of allele frequencies.
Then the \emph{genetic covariance} between samples $a$ and $b$ is
\begin{align*}
    C_{ab}
        &= \frac{1}{L} \sum_j (G_{ja} - P_j) (G_{jb} - P_j) . %  \\
        % &= \frac{1}{L} \left( (G - P \bone^T)^T (G - P \bone^T) \right)_{ab} .
\end{align*}
This can be computed by letting
$\iw_a = \delta_a$ and $\iw_b = \delta_b$
(these weights mark nodes above $a$ and $b$ respectively)
and $\iw_t = \bone / n$
(this weight gives the proportion of samples below each node).
Then, with
$f(\nw_a, \nw_b, \nw_t) = (\nw_a - \nw_t) (\nw_b - \nw_t)$,
we can compute the covariance as
\begin{align*}
    C_{ab} = \frac{1}{2} \bar \site(f, (\iw_a, \iw_b, \iw_t)) .
\end{align*}
This follows, including the factor of two,
because the summary function applied to the ancestral allele
at a site with derived allele frequencies $p_a$, $p_b$, and $p_t$
in sample $a$, sample $b$, and total, respectively, is
$f(1 - p_a, 1 - p_b, 1 - p_t) = f(p_a, p_b, p_t)$.

@@

\plr{Hmph: to actually compute $Cz$ we need one weight vector for each sample, again.
    TODO: look up if you can do Krylov just by computing $y^t C z$ for a reasonable number of vectors.
    We could compute $Cz$ quickly if we could propagate down the tree:
    we'd have $(Cz)_a$ equal to the sum of the weights at all nodes \emph{above} $a$. }

>>>>

Many population genetic summary statistics can be defined using the allele frequency spectrum
of a collection of homologous sequences.
Concretely, write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
-- i.e., the number of alleles whose frequency is $p$ in our sample --
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci, as
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
(Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).
Here, we only consider statistics whose values are invariant to monomorphic sites. Thus, for a sample of $n$
genomes, $0 < p \leq 1-1/n$.

What we do here is a generalization of this idea.
Roughly speaking, to do this we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics
(e.g., averaged in windows).
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular edge in a tree
by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
(b) finding the total weight of all nodes in the subtree inheriting from that edge.
It turns out to be useful to allow more general weights (including negative ones).
Figure \ref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarising them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

@@

We allow vector-valued weights,
i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
so that when summarising we have access to more than one aspect of each subtree.
The requirement that $f(0) = f(\tiw) = 0$ is for consistency,
so that statistics do not depend on portions of the tree either not ancestral to any of the samples or ancestral to all
of them.
% Below, in section XXX, we describe how to efficiently
% maintain a list of the weights of the subtrees inherit from each node
% as we iterate through the tree sequence.

\begin{figure}
    \centering
    \includegraphics{figures/divergence_diagram}
    \caption{
        An example of computing sequence divergence between red nodes and blue nodes
        from seven mutations at distinct sites in a sequence of length 10 using the tree.
        The summary function, $f(a,b) = (a/3)(1-b/2)$ gives the probability that
        an allele found in $a$ of the red nodes and $b$ of the blue nodes is present in
        a randomly chosen red node but not in a randomly chosen blue node.
        \textbf{(a)} Weights are assigned to each subtree by adding together the weights of any children,
        plus the weight of the node itself, if it is a sample.
        \textbf{(b)} In this example, subtree weights $(a,b)$ record how many red nodes ($a$)
        and how many blue nodes ($b$) lie in that subtree.
        \textbf{(c)} The summary function shown is applied to both the weight of the subtree
        and the remaining weight of each mutation,
        so for instance the mutation above the node with weights $(2,1)$
        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
        The final value for divergence is the sum of these values divided by 10,
        the total number of sites.
        \krt{This is a nice fig, but the $2 \times$ and the $/10 sites$ parts may be easy to miss?}
        \label{fig:divergence_diagram}
    }
\end{figure}

@@

Before defining the different classes of statistic,
it will help to establish a little more notation.
A tree sequence efficiently describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
% These trees are highly correlated,
% a fact which allows them to be stored and processed very efficiently,
% as described in \citet{kelleher2018efficient}.
From the tree sequence can be extracted a sequence of trees,
$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome betwen $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance the branch describing a relationship where $u$ is the child of $v$
is associated with $u$.
The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
and is denoted $t_T(u)$ (dropping $T$ if it is clear from context).
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships
can cause the precise set of samples that inherit from that edge to differ across
trees due to recombination.

@@

Perhaps the most natural way to summarise weights on the tree
is simply to examine the values at each node.
This would allow us, for instance, to
count the number of samples that inherit from each node in the tree.
Averaged across trees,
this tells us what proportion of the sample's genomes were inherited from each node.
Motivated by this, we define the
\textbf{node statistic} for node $u$
associated with summary function $f()$ and sample weights $\iw$
to be the sum of $f()$ applied to the weight of the subtree inheriting from node $u$
and to the remaining weight,
averaged across the genome:
\begin{align}
    \node(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k \left( f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right).
\end{align}
The weight $\nw_{T_k}(u)$ is the subtree weight of node $u$ in the tree $T_k$,
and $\tiw - \nw_{T_k}(u)$ is the sum of all weights \emph{not} in the subtree of $u$ in $T_k$.
% i.e., the length of the segment of genome that the tree extends for.
% If this is computed over only a ``window'' of the genome
% then $L_k$ is replaced by the length of the portion of that window
% that $T$ extends for.

@@

We will use weights that count number of samples inheriting from each node
sufficiently often that we give them a special name --
for a subset set $S$ of samples,
the \emph{indicator weights of $S$} are the sample weights $\bone_S$ with
$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.

\begin{example}[Ancestry proportions] \label{ex:ancestry_props}
    If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
    and then $\iw_u / n$ is the proportion of the samples in $S$ that inherit from $u$.
    Therefore, if $f(x) = x / n$,
    then $\nodep(f, \iw)_u$ is the proportion of the genomes of $S$
    that are inherited from (ancestor) $u$.
\end{example}

@@

The summaries we defined above for each node in each tree
extend nicely to genetic variants
-- just as a node weight contains information about which samples inherit from the node and which are not,
so we can summarise patterns of genetic variation by summing up weights of all samples
that carry a given allele.
So, we define \emph{allele weights} to be
the total weight of all sample nodes who have inherited that allele.

@@

If there has been only one mutation at the site,
then $\aw_j(a)$ is equal to the weight of the subtree inheriting from the mutation that produced $a$.
More generally, $\aw$ is not necessarily equal to a subtree weight
but is easily computable from them.

@@

\begin{example}[Nucleotide diversity] \label{ex:site_diversity}
    The nucleotide diversity of a sample is the average number of differences between all haplotypes, which is
    equivalent to the sum of heterozygosity at all positions.
    To calculate this statistic,
    % let $\iw_u = 1$ for $u \in S$ and $\iw_v = 0$ otherwise,
    let $\iw = \bone_S$,
    so that $\nw(u)$ gives the number of nodes in $S$ inheriting from $u$,
    and define
    \begin{align*}
        f(x) = \frac{x (n - x)}{n (n-1)} .
    \end{align*}
    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide diversity in the region between $a$ and $b$.
\end{example}

\begin{example}[Nucleotide divergence] \label{ex:site_divergence}
    Now suppose we want to compute the mean density of nucleotide differences
    \emph{between} two nonoverlapping groups of samples, $S_1$ and $S_2$,
    with $n_1$ and $n_2$ samples, respectively.
    As before,
    let $\iw_j = \bone_{S_j}$,
    so that $\nw_{j}(u)$ gives the number of nodes in $S_j$ inheriting from $u$.
    Then we define
    \begin{align*}
        f(x_1, x_2) = \frac{x_1 (n_2 - x_2)}{n_1 n_2} .
    \end{align*}
    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide divergence between the two groups
    in the region between $a$ and $b$.
\end{example}

@@

This makes it natural to define
a statistic of tree shape by summing these expected contributions across its branches.
We define the \textbf{branch statistic} of a single tree $T$ with span $L$
obtained from summary function $f()$
and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
to be the sum of the area of each branch in each tree
multiplied by the summary function applied to its subtree weight:
\begin{align}
    \branch(f, \iw)_T
    &=
    \frac{1}{L} \sum_{u \in T} t_{u} \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
\end{align}
where $t_{u}$ is the length of the branch ancestral to node $u$ in tree $T$
and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$,
because the sum of all weights in the tree is always $\tiw$.
The value $f(\nw(u))$ is the summary value that would be added to a site statistic
if a single mutation occurred on the branch ancestral to $u$,
and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
so $\branch(f, \iw)_T$ gives the expected contribution of this branch to $\site(f, \iw)$,
per unit of sequence length that the tree $T$ covers.

An example is shown in Figure~\ref{fig:branch_site_diagram},
showing how the $f_4$ site statistic assigns a weight to each mutation
depending on its' frequency in each of the four sample sets,
and how the corresponding branch statistic assigns the same weight to those branches.

@@

\begin{example}[Mean TMRCA] \label{ex:branch_diversity}
    If we take $\iw$ and $f$ exactly as in the ``Nucleotide diversity'' example above,
    then $f(u)$ gives the probability that the branch ancestral to $u$
    lies on the path from two randomly chosen samples from $S$
    on the path up to their most recent common ancestor (MRCA).
    Therefore, $\branch(f, \iw)$ for these choices
    gives the mean total distance in the tree between two samples from $S$,
    averaged across the sequence.
    This is twice the mean TMRCA if the samples are all from the same time.
\end{example}

@@

Thus far we have examined how statistics vary along the genome,
but a tree sequence allows us access to another dimension: time.
Next, we examine how the \emph{time} distribution of admixture signal
in the tree sequence inferred by tsinfer/GEVA.
The branch $f_4(A, B; C, D)$ statistic for four populations $A$, $B$, $C$, and $D$
sums $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and height,
where $p_A$ is the proportion of samples in $A$ that inherit from that node, etcetera.
A branch with a large value is expected to contribute a large amount to the site $f_4$,
which we calculate from data and is used to detect historical connections between populations.
To examine the time signature, we assigned the contribution of each branch
to its terminal node,
and then summed these values across all nodes in 50 logarithmically-spaced bins by time.
We then normalised these to sum to 1 by dividing by the total branch $f_4$ value.

@@

\paragraph{Diploid statistics}
The statistics we develop here are in essence \emph{haploid} statistics.
These tools can be used to compute a statistic about polyploid individuals,
but they do not scale to large numbers of individuals.
For instance, to compute the heterozygosity of a set of samples
-- this is $(1/N) \sum_{i=1}^N 2 p_i (1-p_i)$, where $p_i$ is the derived allele frequency
in the $i^\text{th}$ individual --
one would need to have an entry in the weight vector for \emph{each individual}.
The genotype of a diploid at a site with a single mutation that occurred on the branch ancestral to node $n$
is determined by where that node falls relative to the diploid's two haplotypes in the tree:
if $n$ is on the path from either haplotype to their MRCA, the individual is heterozygous;
if $n$ is on the path from the MRCA to the root, the individual is homozygous derived;
and the individual is homozygous ancestral otherwise.
This suggests the following generalization of our current scheme for diploids:
for a list $(I_1, \ldots, I_N)$ of pairs of nodes,
and two vectors $w^{(1)}$ and $w^{(2)}$ of weights of length $N$,
add weight $w^{(1)}_i$ to each node on the path between the two nodes of $I_i$,
and weight $w^{(2)}_i$ to each node on the path from the MRCA of $I_i$ to the root.
With $k$ such weighting schemes, we would then need the summary function $F$
to take $k \times 2$ values as input.
Efficiently updating these ``diploid'' weights on the nodes should be possible,
but is not a straightforward extension of our current method.

@@

Let $G$ denote the genotype matrix, so that $G_{ja} \in \{0,1\}$
is the genotype of the $a^\text{th}$ individual
at the $j^\text{th}$ site, with `1` denoting the derived state
(we assume here biallelic markers).
Also let $P_j = (1/n) \sum_a G{ja}$ denote the vector of allele frequencies.
Then the \emph{genetic covariance} between samples $a$ and $b$ is
\begin{align*}
    C_{ab}
        &= \frac{1}{L} \sum_j (G_{ja} - P_j) (G_{jb} - P_j) . %  \\
        % &= \frac{1}{L} \left( (G - P \bone^T)^T (G - P \bone^T) \right)_{ab} .
\end{align*}
This can be computed by letting
$\iw_a = \delta_a$ and $\iw_b = \delta_b$
(these weights mark nodes ancestral to $a$ and $b$ respectively)
and $\iw_t = \bone / n$
(this weight gives the proportion of samples inheriting from each node).
Then, with
$f(\nw_a, \nw_b, \nw_t) = (\nw_a - \nw_t) (\nw_b - \nw_t)$,
we can compute the covariance as
\begin{align*}
    C_{ab} = \frac{1}{2} \bar \site(f, (\iw_a, \iw_b, \iw_t)) .
\end{align*}
This follows, including the factor of two,
because the summary function applied to the ancestral allele
at a site with derived allele frequencies $p_a$, $p_b$, and $p_t$
in sample $a$, sample $b$, and total, respectively, is
$f(1 - p_a, 1 - p_b, 1 - p_t) = f(p_a, p_b, p_t)$.

@@

\plr{Hmph: to actually compute $Cz$ we need one weight vector for each sample, again.
    TODO: look up if you can do Krylov just by computing $y^t C z$ for a reasonable number of vectors.
    We could compute $Cz$ quickly if we could propagate down the tree:
    we'd have $(Cz)_a$ equal to the sum of the weights at all nodes \emph{ancestral to} $a$. }
